---
title: "Tutorial for main functions in MonotonicityTest package"
author: "Dylan Huynh"
date: "2025-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

This package provides functions to visualize data and perform monotonicity tests using nonparametric and bootstrap-based methods. The goal is to test if some response variable `Y` is strictly increasing, or decreasing with larger values of some predictor `X`. A detailed explanation of this test is available in Hall and Heckman (2000, The Annals of Statistics, doi:10.1214/aos/1016120363).

The main functions provided are `monotonicity_test` and `create_kernel_plot`. 
The function `monotonicity_test` conducts a nonparametric bootstrap test for monotonicity between two numeric vectors `X` and `Y`. The `create_kernel_plot` provides a visualization of the kernel regression we will do based on a specific kernel bandwidth.

We will use this to analyze the `Wage` dataset from `ISLR` which contains wage data for 3000 male workers. We use this dataset
as we recommend having 300+ points for the kernel estimation being used by the test to be accurate. 

## Preparing Data & Loading Libraries
```{r, warning=FALSE}
# If not installed,
# install.packages("ISLR")
# install.packages("MonotonicityTest")

# Loading Libraries
library(MonotonicityTest)
library(ISLR)

# Loading data
data(Wage)
data <- Wage

head(data)
```

We are particularly interested in two variables,

- **`age`**: Age of the worker
- **`wage`**: Workers raw wage

We would expect `wage` to monotonically increase with the age, and want to analyze this.


## Choosing Kernel Bandwidth

The first thing we need to do is choose an appropriate kernel bandwidth using the `create_kernel_plot` function.

```{r}
X <- data$age
Y <- data$wage

create_kernel_plot(X, Y, bandwidth = seq(1, 5, by=0.5), nrow=3)
```

From here, `bandwidth=3` seems like a fitting choice for the bandwidth.

## Performing Test

We can now use the `monotonocity_test` function to test monotonicity between the `age` and `wage` columns. 
I will set `ncores=4` to speed up the computation and `boot_num=200` for a good balanced between runtime and precision. I will also set the random seed for reproducability.

```{r}
seed <- 42
set.seed(seed)

bandwidth <- 3
ncores <- 4
boot_num <- 20

res <- monotonicity_test(
  X, 
  Y,
  ncores = ncores,
  boot_num = boot_num,
  bandwidth = bandwidth,
  negative= FALSE,
  seed = seed,
  check_m = TRUE  # Run test for different values of m, has no impact on runtime performance
)

print(res)
```

From the output we get our p-value, our test-statistic, and the bootstrap distribution of the test-statistic under the null, that the function is monotonic. If this p-value were less than our chosen $\alpha$ we would reject the null in favor of the alternative hypothesis, that the function is non-monotonic. We can further call `summary` and `plot` to evaluate the result
in further detail.

```{r}
summary(res)
plot(res)
```
